Managing variability in software systems

### 2021-05-02 Paper.tex
origins of variability

Decision modeling has its earliest documented approach in the Synthesis method. \cite{b6} This defines it as \textit{a set of decisions that are adequate to distinguish among the members of an application engineering product family and to guide adaptation of application
	engineering work products}. \cite{b7}\\

\section{Origin of variability (1/2 page) NH}
We introduce why there is variability in the modern software engineering process. So that the reader knows how the problems arrise to have a baseline for presenting solutions for said problems.

\section{Variability modeling (1 page) JR}
As mentioned there are two main approaches for managing variability. For each of which we will discuss the advantages and disadvantages and create a broad overview over the different modeling approaches.\\\\
\textbf{Feature modeling}\\
Features in the context of feature modeling are behaviours, characteristics or functionalities which are observable by the user of the software system. They can be used to define commonalities and differences between systems and their variants, as well as determine mandatory, optional or alternative aspects of them\cite{b8}. A broadly used approach of displaying features and their relations is hierarchically structured diagram which is called a features diagrams.\\\\
A big advantage of feature models, especially in the shape of feature diagrams, is a possible communication enhancement between stakeholders. This can be achieved by providing a overview of all functionalities which are offered to the user by the system. Due to the nature of the model regarding structure and due to the fact that it displays if and which planned features are already provided by another system or system family, this may improve the cost and time estimation for developing the software system. \\\\
The essence of feature modeling is the splitting of features into distinct objects which have a certain relation to each other. There are several types of features mentioned in \cite{b8}, named \textit{mandatory, optional and alternative features}. Generally speaking common attributes of systems are described by mandatory features, while alternative or optional features define the variability in feature models.\cite{b9}\\\\ 
If a feature is mandatory it has to be selected if its parent feature is selected, the type of the parent feature does not matter. A optional feature can be selected freely but there may be constraints from composition rules (e.g. a feature can only be choosen as part of a feature sets). Alternative features are can be pictured as a choice box, so one can choose \textit{one} item from a feature set.\\\\
This is the approach which is used in the original FODA paper, which was extended with the introduction of cardinalities for alternative features. \cite{b10} This expansion grants the ability to create minimum and maximum constraints for choosing alternative features from a feature set, which helps with creating complex models.\\\\
In Fig. 1, one can see a small feature diagram of a car, it consists of one root node, that describes the domain we are modeling, in this case the car. The previous elaborated mandatory features are the engine and seats. Furthermore there is the option to install a cruise control and the option for a turbocharger these features are completely independet and optional. Last but not least one can choose between normal leather seats or alcantara ones but not both. Which this simplified model of a car we have all three types of features modeled

@@

\section{Variability patterns}
Sums up the typical patterns how variability can be introduced.
(1/2 page, maybe merge with section above) NH

\section{???Variability planning???}
Gives an overview of how someone can plan for varability and how to manage it.

>>>>

Decision modeling has its earliest documented approach in the Synthesis method. \cite{b6} This defines it as \textit{a set of decisions that are adequate to distinguish among the members of an application engineering product family and to guide adaptation of application
	engineering work products}. [7 p. 104]\\

All in all there exists various literature on the different scopes variability management encompasses and we try to capture the main points of several of them in this paper.

\section{Origin of variability (1/2 page) NH}
Variability is the ability of software to be changed for specific context. That encompasses extension, customization and reconfiguration. It thus defines the parts of the software that remain variable and are not fully specified yet. \cite{b4}\\\\
There are two main cases where one needs to capture different alternatives within an architecture:\\
\begin{enumerate}
	\item At the time of design there may exist various alternatives and if a decision cannot be made yet because of lack of information, it will be deferred. Therefore one must design the system to be able to handle all viable alternatives.\\
	\item There already exists an product line which contains several different alternatives for given parts of software, which can be chosen depending on the needs of each individual customer.\\
\end{enumerate}
While these situations differ greatly in reality, in the context of variability management they are really the same. That is, before a decision is made on which alternative is going to be chosen, one does not know the outcome and needs to be able to handle all possible ones.\cite{b3}\\

There are also several reasons why such variations have to exist in the first place \cite{b3}:
\begin{enumerate}
	\item \textit{Variation in function.} Not every product supports the same functions. Take as example a modern smartphone, some might support fingerprint authentication, some might support facial recognition authentication and some might support none of those.\\
	\item \textit{Variation in data.} Products might vary in the definitions of data structures. For example a software managing planned repairs of industrial machines might need store additional information for some companies like an SAP Id in case these planned repairs are also concurrently managed in SAP, while some other companies might not need that information.\\
	\item \textit{Variation in technology.} The technology different products run on may vary. Some software systems for example need to run on Linux while others need to support Windows. Or for some a simple MySQL database might suffice while others need the SQL Server Enterprise edition because the amount of data might vary significantly.\\
	\item \textit{Variation in environment.} The environment software exists in might vary. Meaning also the way in which it interacts with its environment. \\
\end{enumerate}
These, at least, are the four most common origins of variation. Of course some more do exist, but these are beyond the scope of this document.\\
\section{Variability modeling (1 page) JR}
As mentioned there are two main approaches for managing variability. For each of which we will discuss the advantages and disadvantages and create a broad overview over the different modeling approaches.\\\\
\textbf{Feature modeling}\\
Features in the context of feature modeling are behaviours, characteristics or functionalities which are observable by the user of the software system. They can be used to define commonalities and differences between systems and their variants, as well as determine mandatory, optional or alternative aspects of them\cite{b8}. A broadly used approach of displaying features and their relations is hierarchically structured diagram which is called a features diagrams.\\\\
A big advantage of feature models, especially in the shape of feature diagrams, is a possible communication enhancement between stakeholders. This can be achieved by providing a overview of all functionalities which are offered to the user by the system. Due to the nature of the model regarding structure and due to the fact that it displays if and which planned features are already provided by another system or system family, this may improve the cost and time estimation for developing the software system. \\\\
The essence of feature modeling is the splitting of features into distinct objects which have a certain relation to each other. There are several types of features mentioned in \cite{b8}, named \textit{mandatory, optional and alternative features}. Generally speaking common attributes of systems are described by mandatory features, while alternative or optional features define the variability in feature models.\cite{b9}\\\\ 
If a feature is mandatory it has to be selected if its parent feature is selected, the type of the parent feature does not matter. A optional feature can be selected freely but there may be constraints from composition rules (e.g. a feature can only be choosen as part of a feature sets). Alternative features are can be pictured as a choice box, so one can choose \textit{one} item from a feature set.\\\\
This is the approach which is used in the original FODA paper, which was extended with the introduction of cardinalities for alternative features. \cite{b10} This expansion grants the ability to create minimum and maximum constraints for choosing alternative features from a feature set, which helps with creating complex models.\\\\
In Fig. 1, one can see a small feature diagram of a car, it consists of one root node, that describes the domain we are modeling, in this case the car. The previous elaborated mandatory features are the engine and seats. Furthermore there is the option to install a cruise control and the option for a turbocharger these features are completely independet and optional. Last but not least one can choose between normal leather seats or alcantara ones but not both. Which this simplified model of a car we have all three types of features modeled

@@

\section{Variability patterns}
Variability can be introduced into a system via different mechanisms. These mechanisms strive to bind a given system to one out of many alternatives at different binding times. Binding times describe when exactly these mechanisms stop deffering the decision and make the choice for any of the variations, as illustrated in Figure 1.\cite{b2}

\begin{figure}[h!]
	\includegraphics[width=\columnwidth]{binding_times}
	\caption{Binding Times \cite{b2}}
\end{figure}

Further explanation of these binding times can be found in the subsequent paragraphs:\\
\begin{itemize}
	\item \textit{Product Architecture Derivation.}
\end{itemize}
\section{???Variability planning???}
Gives an overview of how someone can plan for varability and how to manage it.
